Adaptive Sequential Model Selection

### 2015-06-18 paper.tex
done with reorg

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\tL}{\widetilde{L}}
\newcommand{\hk}{\hat{k}}
\newcommand{\hr}{\hat{r}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}

@@

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. Our goal is to examine the data and output a {\em stopping rule}, i.e. an estimator $\hk(Y)$ of $k_0$. We consider $M_k$ to be ``rejected'' if $k < \hat k$, and ``accepted'' otherwise. Thus $\hk$ is the number of models we rejected, while $k_0$ is the number that we should have rejected. The number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$.

We will break the problem into two constituent subproblems. First, Section~\ref{sec:singleStep} will discuss how to obtain selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{0,k}:\; F\in M_{k-1}\quad \text{ vs. } \quad 
H_{1,k}:\; F\in M_\infty\setminus M_{k-1},
\]
adjusting for the fact that the models are chosen adaptively. In most of our development we will make few assumptions about what form the candidate models take. Still, our results are only useful in cases where we can actually construct single-step $p$-values. At present, this largely restricts our applications to exponential family models.

Second, Section~\ref{sec:sequential} focuses on the sequential nature of the problem. \citet{gsell2013sequential} and \citet{li2015accumulation} propose stopping rules that operate on the sequence $p_{[d]}$. These stopping rules provably control type I error rates like the familywise error rate $\P(V>0)$ or the false discovery rate $\E[V/(\hk \vee 1)]$, provided that the null $p$-values are mutually independent and independent of the non-null $p$-values. We will discuss general conditions under which single-step $p$-values are independent.

@@

Most of our examples will focus on the important and familiar problem of adaptively selecting a set of predictors in linear regression models. In this problem, we observe a random response $Y\in \R^n$ as well as a fixed design matrix $X\in \R^{n \times p}$, whose columns correspond to candidate predictors. For each {\em active set} $E \sub [p]$, there is a corresponding candidate model
\[
M(E):\; Y \sim \cN( X_E\beta, \sigma^2 I_n),
\]
which is a subset of the {\em full model}
\[
M_\infty:\; Y \sim \cN(X\beta, \sigma^2I_n).
\]
The error variance $\sigma^2$ may be known or unknown depending on the context.

@@

All of the specific examples considered in this paper will have a common form generalizing the linear regression problem above. Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^{\cJ}$:
\[
M_\infty = \{F_\theta:\; \theta \in \Theta\}.
\]
For any subset $E\sub \cJ$ define the sparse submodel with active coefficients $E$ as follows:
\[
\Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
\quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
\]

@@

\paragraph{Forward Stepwise}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. The {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed $E_0$, then at step $k=1,\ldots,d$, we define
\begin{align}\label{eq:forwardDef_start}
j_k &= \argmax_j \;\;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\} \\
E_k &= E_{k-1} \cup \{j_k\}\\\label{eq:forwardDef_end}
M_k &= M(E_k).
\end{align}
That is, at each step we add to the model the one variable that would most increase the likelihood. In the case of linear regression, this amounts to adding the variable that maximally reduces the residual sum of squares.

@@

@@

@@

Having discussed methods for constructing single-step $p$-values, we turn now to the question of using them to construct a stopping rule $\hat k$. Specifically, we will consider stopping rules that depend on the sequence $p_1,\ldots,p_d$ of $p$-values and must always reject an initial block of corresponding hypotheses.

Section~\ref{sec:randomHyp} outlines several proposals for ordered testing of hypotheses and discusses what error rates they control and under what conditions. We focus on proposals from \citet{gsell2013sequential} and \citet{li2015accumulation}.

These two papers both assume that the hypotheses $H_1, \ldots, H_d$ are fixed, and in particular that $k_0$ is fixed. In our problem, the model path is random and thus so is $k_0$. As we will see in Section~\ref{sec:randomHyp}, \WFcomment{... that this is not a problem}.

While StrongStop, ForwardStop, and the accumulation tests tend to give more powerful stopping rules, they require independence among the $p$-values. Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which selective $p$-values are mutually independent. We will see that, typically, selected-model $p$-values are sequentially independent while saturated-model $p$-values are not.

\subsection{Ordered Testing of Random Hypotheses}\label{sec:randomHyp}

\subsubsection{BasicStop}
The most obvious procedure is simply to reject at each step until the first time that $p_k > \alpha$, which we can formalize as
\[
\hk_B(Y) = \min\left\{k \in \{1,\ldots,m\} :\;
  p_k > \alpha\right\} - 1
\]
We will call this procedure {\em BasicStop}. \WFcomment{Is there a better name for this? Can we find it in the literature?}

Selectively valid single-step $p$-values do not necessarily guarantee FWER control when $k_0$ is random. In that case, we need a bit more, but it is sufficient that we have type I error control for $p_k$ conditional on $k_0=k-1$; i.e., not just conditional on $M_{k-1}$ being a correct model, but given that $M_{k-1}$ is the {\em first} correct model.

\citet{gsell2013sequential} propose another stopping rule that controls the model-wise FWER, which they called {\em strong stop}. Define
\[
  \hk_{S}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    \exp\left(\sum_{i=k}^d \frac{\log p_i}{i}\right) 
    \leq \frac{\alpha k}{d}\right\}
\]
Even if $p_k$ is a little larger than $\alpha$, say $\alpha=0.05$ and $p_k=0.06$, StrongStop can still reject $M_{k-1}$ if the next three or four $p$-values are also relatively small.

\citet{gsell2013sequential} show that if the completion index $k_0$ is fixed, then $\hk_S$ controls the FWER at level $\alpha$, as long as the null $p$-values are independent given the non-null ones. Specifically, they require that
\begin{equation}\label{eq:indepCond_fixed_k0}
\P(p_{k_0+1} \leq \alpha_{k_0+1}, \ldots, p_d \leq \alpha_d
\mid p_1, \ldots, p_{k_0}) \leq \prod_{i=k_0+1}^d \alpha_i
\end{equation}

For random $k_0$, it is easy to see that the $p$-values simply need to satisfy~\eqref{eq:indepCond_fixed_k0} conditional on the value of $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_{k_0}, \; k_0 = k) \leq \prod_{i=k+1}^d \alpha_i
\end{equation}
If~\eqref{eq:indepCond_random_k0} holds, then strong stop controls the probability of $V>0$ conditionally on $k_0$, and thus also marginally.

\citet{gsell2013sequential} propose another stopping rule, {\em forward stop}, that controls the model-wise FDR:
\[
  \hk_{F}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
If $p_k$ is uniform then  $-\log(1-p_k)$ is an exponential random variable. If all of the null $p$-values are uniform and the others are zero, then only the null ones contribute to the sum, and
\[
\widehat{\text{FDR}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i)
\]
is a Gamma random variable with mean $V/k$ and variance $V/k^2$. We stop the last time the estimate of FDR is less than $\alpha$. Forward stop requires the same conditional independence condition~\eqref{eq:indepCond_random_k0} as strong stop does.

\subsection{$p$-Values and Independence}\label{sec:pValsIndep}

Recall that $p_k$ is a valid selective $p$-value for $M_{k-1}\sub M_\infty$ if, for all $F\in M_\infty$,
\[
\P_F(p_k \leq \alpha \mid M_{k-1},\; F\in M_{k-1}) \leqAS \alpha.
\]
Valid single-step selective $p$-values do not generally result in $p$-values that are independent from one step to the next.

For example, saturated-model $p$-values are generically non-independent. Continuing with Example~\ref{ex:bivariate}, Table~\ref{tab:bv_twoWayTable} shows a two-way contingency table for the saturated-model $p$-values $(p_1(Y), p_2(Y))$, binned into cells of height and width 0.2, simulated under the global null $\mu=0$. Because $k_0=0$, both $p$-values are uniform by construction, but the $p$-values are strongly dependent, with correlation $-48\%$. By contrast, we will see that the selected-model $p$-values $(p_1,p_2)$ are independent under the global null, a consequence of Proposition~\ref{prop:selectiveModel}.

@@

First, note that $p_2$ is a selective $p$-value for comparing the null with only variable 1,
\[
M_1:\; Y \sim \cN\left(\binom{\mu_1}{0}, I_2\right),
\]
against the alternative with variables 1 and 2:
\[
M_2:\; Y \sim \cN\left(\mu, I_2\right), \text{ with } 
\mu_2 \neq 0.
\]
To eliminate the nuisance parameter $\mu_1$, the test conditions on $Y_1$. Thus, $p_2$ is uniform given $Y_1$ on $A$. Second, note that on $A$, $p_1$ is a function only of $Y_1$. Thus, $(p_1,p_2)$ are independent uniform variables given that variable 1 is chosen first. By a similar argument, they would also be independent uniforms if $Y_2$ were chosen first; thus, $(p_1, p_2)$ are marginally uniform under the global null.

We now introduce a simple sufficient condition to verify that a given procedure yields independent $p$-values. We say that a filtration $\sF_{[d]}$ with $M_{[k]} \in \sF_k$ {\em separates} the $p$-values $p_{[d]}$ if 
\begin{enumerate}
\item $p_k(Y)$ is conservative given $\sF_{k-1}$ 
  when $F\in M_{k-1}$, and
\item $p_k(Y)$ is measurable with respect to $\sF_k$.
\end{enumerate}
If we think of $\sF_k$ representing information available at step $k$, then the first condition means that $p_k$ excludes whatever evidence may have accrued against the null by step $k-1$, and the second means that information revealed after step $k$ plays no role in determining $p_k$. In that sense, $\sF_k$ forms a ``wall of separation'' between the $p$-values up to $k$ and the ones after $k$. As a result, separated null $p$-values are independent.

>>>>

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\tL}{\widetilde{L}}
\newcommand{\hk}{\hat{k}}
\newcommand{\hr}{\hat{r}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}
\newcommand{\Err}{\mathcal{E}}

@@

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. Our goal is to examine the data and output a {\em stopping rule}, i.e. an estimator $\hk(Y)$ of $k_0$. We consider $M_k$ to be ``rejected'' if $k < \hat k$, and ``accepted'' otherwise. Thus $\hk$ is the number of models we rejected, while $k_0$ is the number that we should have rejected.

The number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Depending on the context we might want to control the familywise error rate (FWER) $\P(V>0)$, the false discovery rate (FDR) $\E[V/(\hk \vee 1)]$, or another error rate given by the expectation of some loss function $g$:
\begin{equation}\label{eq:errRate}
\Err_F(\hk(\cdot), g) = \E_F\left[ g\left(\hk(Y), k_0(Y)\right)\right]
\end{equation}

We will assume that for each candidate model $M$ there is a minimal sufficient statistic $T(Y; \,M)$, and write
\[
T_k(Y) = T(Y; M_k(Y)).
\]

We will break the problem into two constituent subproblems. First, Section~\ref{sec:singleStep} will discuss how to obtain selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{k}:\; F\in M_{k-1}(Y)
\]
against the alternative that $F\in M_\infty\setminus M_{k-1}$. The $p$-values must account for the fact that the models are chosen adaptively. In most of our development we will make few assumptions about what form the candidate models take. Still, our results are only useful in cases where we can actually construct single-step $p$-values. At present, this largely restricts our applications to exponential family models.

Second, Section~\ref{sec:sequential} focuses on the sequential nature of the problem. \citet{gsell2013sequential} and \citet{li2015accumulation} propose stopping rules that operate on the sequence $p_{[d]}$. These stopping rules provably control certain error rates, provided that the null $p$-values are mutually independent and independent of the non-null $p$-values. We will discuss general conditions under which single-step $p$-values are independent.

@@

Most of our examples will focus on the important and familiar problem of adaptively selecting a set of predictors in linear regression models. In this problem, we observe a random response $Y\in \R^n$ as well as a fixed design matrix $X\in \R^{n \times p}$, whose columns correspond to candidate predictors. For each {\em active set} $E \sub [p]$, there is a corresponding candidate model
\[
M(E):\; Y \sim \cN( X_E\beta, \sigma^2 I_n),
\]
which is a subset of the {\em full model}
\[
M_\infty:\; Y \sim \cN(X\beta, \sigma^2I_n).
\]
The error variance $\sigma^2$ may be known or unknown depending on the context. The complete sufficient statistic for $M(E)$ is $X_E'Y$ if $\sigma^2$ is known, or else $\left(X_E'Y, \|Y\|^2\right)$.

@@

All of the specific examples considered in this paper will have a common form generalizing the linear regression problem above. Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^{\cJ}$:
\[
M_\infty = \{F_\theta:\; \theta \in \Theta\}.
\]
For any subset $E\sub \cJ$ define the sparse submodel with active coefficients $E$ as follows:
\[
\Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
\quad M(E) = \{F_\theta:\; \theta\in \Theta(E)\}.
\]

@@

\paragraph{Forward Stepwise Paths}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. The {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed $E_0$, then at step $k=1,\ldots,d$, we define
\begin{align}\label{eq:forwardDef_start}
j_k &= \argmax_j \;\;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\} \\
E_k &= E_{k-1} \cup \{j_k\}\\\label{eq:forwardDef_end}
M_k &= M(E_k).
\end{align}
That is, at each step we add to the model the one variable that would most increase the likelihood. In the case of linear regression, this amounts to adding the variable that maximally reduces the residual sum of squares.

@@

\WFcomment{Shorten this section, move most of it to Discussion}

@@

An ordered hypothesis testing procedure takes in a sequence of $p$-values $p_1, \ldots, p_d$ for null hypotheses $H_{1}, \ldots, H_{d}$, and outputs a decision $\hk$ to reject the initial block $H_{1}, \ldots, H_{\hk}$ and accept the remaining hypotheses. We will assume throughout that the hypotheses are nested with $H_{k-1} \sub H_k$. Section~\ref{sec:orderedProposals} outlines several proposals for ordered testing and discusses what error rates they control and under what conditions.

We focus on proposals from \citet{gsell2013sequential} and \citet{li2015accumulation}, all of which assume that the null hypotheses $H_{1}, \ldots, H_{d}$ are fixed, and that the null $p$-values are independent of each other and of the non-null $p$-values. Specifically, they assume that
\begin{equation}\label{eq:indepCond_fixed_k0}
\P(p_{k_0+1} \leq \alpha_{k_0+1}, \ldots, p_d \leq \alpha_d
\mid p_1, \ldots, p_{k_0}) = \prod_{i=k_0+1}^d \alpha_i,
\end{equation}
where $k_0$ is fixed.

In our problem the null hypotheses at each step are random and thus so is the completion index $k_0(Y)$. We will thus require {\em conditional} independence of null $p$-values given $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_{k_0}, \; k_0 = k) = \prod_{i=k+1}^d \alpha_i
\end{equation}

The following proposition shows that~\eqref{eq:indepCond_random_k0} is the correct generalization of~\eqref{eq:indepCond_fixed_k0} to the random-hypothesis case.
\begin{proposition}
  Let $\hk$ be a stopping rule that operates on sequences
  $p_{[d]}(Y)$ of $p$-values for ordered hypotheses $H_{[d]}(Y)$. 
  For some function $g$, let 
  \[
  \Err(\hk,g) = \E\left[ g\left(\hk(Y), k_0(Y)\right)\right],
  \]

  Suppose that the rule $\hk$ controls $\Err$ at $\alpha$ under
  independence when $H_{[d]}$ are fixed. That is, suppose that
  whenever $H_{[d]}$ are fixed and $p_{[d]}$
  satisfy~\eqref{eq:indepCond_fixed_k0}, we have
  \[
  \Err(\hk, g) \leq \alpha
  \]

  Then, $\hk$ controls $\Err$ at $\alpha$
  whenever $p_{[d]}$ satisfy~\eqref{eq:indepCond_random_k0}.
\end{proposition}
\begin{proof}
  For any fixed $k_0$ and $q_1,\ldots, q_{k_0}$, we can construct a
  problem with $\tilde p_k \sim \delta_{q_k}$ for $k\leq k_0$ and 
  $\tilde p_k \sim \text{Unif}[0,1]$ for $k > k_0$. Then, for the
  fixed hypotheses 
  $\widetilde H_k:\; \tilde p_k \sim \text{Unif}[0,1]$, we have
  \[
  f(k_0, q_{[k_0]}) 
  = \E\left[ g\left(\hk(\tilde p_1, \ldots, \tilde p_d),
      k_0\right) \right],
  \]
  which is less than or equal to $\alpha$ by assumption.

  But then, by the condition~\eqref{eq:indepCond_random_k0}, we have
  \begin{align}
    \Err(\hk,g) &\;=\; \E\left[ \;\E\left[ g\left(\hk(p_1,\ldots, p_d), k_0(Y)\right) \; \mid \; k_0(Y), \; p_{[k_0]}(Y)\right] \; \right]\\
    &\;\leq\; \E\left[f(k_0, p_{[k_0]})\right] \;\leq\; \alpha.
  \end{align}
\end{proof}

In real applications, null $p$-values may be conservative instead of exactly uniform. We can relax~\eqref{eq:indepCond_random_k0} to require only that the joint distribution of null $p$-values is stochastically larger than uniform:
\begin{equation}\label{eq:stochLargeCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_{k_0}, \; k_0 = k) \leq \prod_{i=k+1}^d \alpha_i
\end{equation}
If the null $p$-values are marginally uniform, then the conditions~\eqref{eq:indepCond_random_k0} and~\eqref{eq:stochLargeCond_random_k0} are equivalent. For most stopping rules $\hk$, the weaker condition~\eqref{eq:stochLargeCond_random_k0} is enough to guarantee control of whatever error rate holds under~\eqref{eq:indepCond_random_k0}. For the sake of brevity, we will say that $p$-value sequences satisfying~\eqref{eq:indepCond_random_k0} are {\em independent on nulls}.

Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which~\eqref{eq:stochLargeCond_random_k0} is satisfied.

\subsection{Stopping Rules for Ordered Testing}\label{sec:orderedProposals}

\begin{comment}
\subsubsection{BasicStop}
The most obvious procedure is simply to reject at each step until the first time that $p_k > \alpha$, which we can formalize as
\[
\hk_B(Y) = \min\left\{k \in \{1,\ldots,m\} :\;
  p_k > \alpha\right\} - 1
\]
We will call this procedure {\em BasicStop}. \WFcomment{Is there a better name for this? Can we find it in the literature?}

Selectively valid single-step $p$-values do not necessarily guarantee FWER control when $k_0$ is random. In that case, we need a bit more, but it is sufficient that we have type I error control for $p_k$ conditional on $k_0=k-1$; i.e., not just conditional on $M_{k-1}$ being a correct model, but given that $M_{k-1}$ is the {\em first} correct model.
\end{comment}

\citet{gsell2013sequential} propose another stopping rule that controls the FWER, which they called {\em strong stop}. Define
\[
  \hk_{S}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    \exp\left(\sum_{i=k}^d \frac{\log p_i}{i}\right) 
    \leq \frac{\alpha k}{d}\right\}
\]
Even if $p_k$ is a little larger than $\alpha$, say $\alpha=0.05$ and $p_k=0.06$, StrongStop can still reject $M_{k-1}$ if the next three or four $p$-values are also relatively small.

\citet{gsell2013sequential} show that if the completion index $k_0$ is fixed, then $\hk_S$ controls the FWER at level $\alpha$, as long as the null $p$-values are independent given the non-null ones.

\citet{gsell2013sequential} propose another stopping rule, {\em forward stop}, that controls the FDR:
\[
  \hk_{F}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
If $p_k$ is uniform then  $-\log(1-p_k)$ is an exponential random variable. If all of the null $p$-values are uniform and the others are zero, then only the null ones contribute to the sum, and
\[
\widehat{\text{FDR}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i)
\]
is a Gamma random variable with mean $V/k$ and variance $V/k^2$. We stop the last time the estimate of FDR is less than $\alpha$.

\subsection{Conditions for Independence}\label{sec:pValsIndep}

The single-step $p$-value $p_k$ is selectively valid for testing $H_k:\; F \in M_{k-1}(Y)$ against $M_\infty \setminus M_{k-1}$ if, for all $F\in M_\infty$,
\[
\P_F(p_k \leq \alpha \mid M_{k-1},\; F\in M_{k-1}) \leqAS \alpha.
\]
Valid single-step selective $p$-values are not necessarily independent on nulls, but there is a simple sufficient condition under which independence on nulls is guaranteed.

Consider any filtration
\[
\sF_0 \sub \sF_1 \sub \cdots \sub \sF_d
\]
for which $M_{[k]}$ is measurable with respect to $\sF_k$. We say that $\sF_{[d]}$ {\em separates} the $p$-values $p_{[d]}$ if 
\begin{enumerate}
\item $p_k(Y)$ is conservative given $\sF_{k-1}$ 
  when $F\in M_{k-1}(Y)$, and
\item $p_k(Y)$ is measurable with respect to $\sF_k$.
\end{enumerate}
If $p_k$ is exact, the first requirement is equivalent to requiring that $p_k$ is independent of $\sF_{k-1}$ on the event $F\in M_{k-1}(Y)$.

If we think of $\sF_k$ as representing information available at step $k$, then the first condition means that $p_k$ ``excludes'' whatever evidence may have accrued against the null by step $k-1$, and the second means that any information revealed after step $k$ plays no role in determining $p_k$. In that sense, $\sF_k$ forms a ``wall of separation'' between the $p$-values up to $k$ and the ones after $k$. As a result, separated $p$-values are independent on nulls.

  If the $p$-values are exact then $p_{k+1}, \ldots, p_d$ are
  independent and uniform given $\sF_{k_0}$ on the event $\{k_0=k\}$.
  If they are conservative, then for all
  $\alpha_{k+1},\ldots,\alpha_d \in [0,1]$,
  \begin{equation}\label{eq:jointConserv}
  \P_F\left(p_{k+1}\leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d\;
    \mid\; k_0 \leq k, \; \sF_k\right) \leqAS \prod_{i=k+1}^d
  \alpha_i.
  \end{equation}
\end{proposition}

\subsubsection{The Sufficient Filtration}\label{sec:suffFilt}

We will concern ourselves mainly with the {\em sufficient filtration} for the path $M_{[k]}$, defined as
\[
\sF_k^{\text{suff}} = \sF(M_{[k]}, T_k),
\]
We will suppress the superscript and use $\sF_k$ to mean $\sF_k^{\text{suff}}$ for the remainder of the article.

The sufficient filtration separates $p_{[d]}$ if and only if 
\begin{enumerate}
\item $p_k$ is conservative given $M_{[k-1]}$ and $T_{k-1}$, and
\item $p_k$ is a function of $M_{[k]}$ and $T_k$.
\end{enumerate}

To be a valid selective $p$-value for a single step, $p_k$ only needs to be conservative conditional on $M_{k-1}$. In general, it is more stringent to additionally require that $p_k$ must also condition on $T_{k-1}$ as well as the entire subpath $M_{[k-1]}$. In most cases of interest, however, the ``extra'' conditioning is redundant, for two reasons.

First, most selective tests already condition on $T_{k-1}$ in order to eliminate nuisance parameters. Second, as we will show in Section~\ref{sec:SSP}, for most path algorithms of interest we have
\begin{equation}\label{eq:SSP}
\sF(M_k, \, T_k) = \sF(M_{[k]}, \, T_k).
\end{equation}
Thus, most single-step selective $p$-values we would contemplate using are already exact conditional on $\sF_{k-1}$.

The second requirement, that $p_k$ be $\sF_k$-measurable, has more bite. For example, we will see in Section~\ref{sec:singleStep} that it excludes saturated-model tests in linear regression. By contrast, any UMPU selective test for $M_{k-1}$ against $M_k\setminus M_{k-1}$ is typically measurable with respect to $\sF_k$.

\subsubsection{The Subpath Sufficiency Principle}\label{sec:SSP}

We say that a path algorithm $M_{[d]}(\cdot)$ satisfies the {\em subpath sufficiency principle} (SSP) if~\eqref{eq:SSP} holds for every $k$. Equation~\eqref{eq:SSP} means that if we know $M_k(Y) = M$, then we can compute the entire subpath of models $M_{[k]}$ knowing only $T(Y, \, M)$, the sufficient statistic of model $M$. For example, given that four steps of forward stepwise linear regression yield the active set $E_4 = \{1,5,7,11\}$, the order in which these variables were included can be determined from $T_4 = X_{E_4}'Y$.

For the generic sparse parametric model of Section~\ref{sec:genSparse}, the forward stepwise path satisfies the SSP, as do all ever-active regularization paths. 
Define the restricted log-likelihood
\begin{equation}
\ell_E(\theta; Y) = \left\{\begin{matrix} 
    \ell(\theta; \;Y) & \theta \in \Theta(E)\\ 
    -\infty  & \mathrm{ otherwise.}\end{matrix}\right. 
\end{equation}

\begin{proposition}[Forward Stepwise Satisfies SSP]\label{prop:forwardSSP}
  The forward stepwise path as defined in
  (\ref{eq:forwardDef_start}--\ref{eq:forwardDef_end}) satisfies the 
  sub-path sufficiency principle.
\end{proposition}
\begin{proof}
  For some fixed step $k$ and active set $E$, 
  let $A$ denote the event $\{M_k = M(E)\}$. 
  On $A$, the restricted likelihood $\ell_E$ is proportional to a
  function depending only on $T_k = T(Y; \;M(E))$:
  \[
  \ell_E(\theta; Y) = \ell_E^{T_k}(\theta; T_k) 
  + \ell_E^{Y \mid T_k}(Y \mid T_k).
  \]
  The second term does not depend on $\theta$ 
  because $T(Y; \;M(E))$ is sufficient for $M(E)$.

Equation~\eqref{eq:onlyT} shows that $j_1,\ldots, j_{k-1}$ all depend on $Y$ only through $T_k(Y)$. As a result, it also follows that the enitre sequence $M_{[k]}(Y)$ depends only on $T_k(Y)$.
\end{proof}

\begin{proposition}[Regularized Likelihood Paths Satisfy SSP]\label{prop:regPathSSP}
Ever-active regularized likelihood paths as defined in 
(\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the
sub-path sufficiency principle.
\end{proposition}

\begin{proof}
  Again, for fixed $k$ and $E$ let $A$ denote the event $\{M_k = M(E)\}$ for some fixed $E$, and define $T_k$ as in the proof of Proposition~\ref{prop:forwardSSP}. If we define
\[
\hat\theta^{(E,r)} = \argmin_{\theta\in\Theta(E)} -\ell_E^{T_k}(\theta; T(Y)) + P_r(\theta),
\]
then 
\[
\hat\theta^{(E,r)} \eqAS \hat\theta^{r} \text{ on } 
A \cap 1_{\{r \leq R_k\}}.
\]
But $\hat\theta^{(E,r)}$ depends only on $T_k(Y)$. Therefore, on $A$, we can reconstruct the entire path of solutions once we know $T_k(Y)$.
\end{proof}

Collecting together the threads of this section, we can arrive at our main result: a fairly general sufficient condition for a $p$-value sequence to be independent on nulls.
\begin{theorem}[Sufficient Condition for Independence on Nulls] \label{thm:suffCond}
Assume that $T(Y; M)$ is a complete sufficient statistic for each candidate model $M$, that $p_k$ are exact selective $p$-values for $M_{k-1}$, and that $M_{[d]}(\cdot)$ satisfies the SSP.

If $p_k$ is $\sF_k$-measurable for each $k$, then the $p$-value sequence $p_{[d]}(Y)$ is independent on nulls.
\end{theorem}

\begin{proof}
We apply the definition of completeness to the function 
\[
h(T(Y; \, M)) \;=\; 
\P(p_k(Y) \leq \alpha \mid T(Y; \, M), \, M_{k-1}(Y) = M) - \alpha
\]
If $p_k$ is exact given $M_{k-1} = M$, then 
\[
\E_F[h(T_{k-1}) \mid M_{k-1}=M] \;=\; 0
\] 
for every $F\in M$. Therefore, we must have $h(T_{k-1})\eqAS 0$, so $p_k$ is independent of $\sF(M_k, T_{k-1})$. Because $M_{[d]}$ satisfies the SSP, $p_k$ is also independent of $\sF_{k-1} = \sF(M_{[k-1]}, T_{k-1})$.

If $p_k$ is also $\sF_k$-measurable, then the sufficient filtration separates $p_{[d]}$, implying that $p_{[d]}$ is independent on nulls.
\end{proof}

@@

Saturated-model $p$-values are generically non-independent. Continuing with Example~\ref{ex:bivariate}, Table~\ref{tab:bv_twoWayTable} shows a two-way contingency table for the saturated-model $p$-values $(p_1(Y), p_2(Y))$, binned into cells of height and width 0.2, simulated under the global null $\mu=0$. Because $k_0=0$, both $p$-values are uniform by construction, but the $p$-values are strongly dependent, with correlation $-48\%$. By contrast, the selected-model $p$-values $(p_1,p_2)$ are independent under the global null, a consequence of Theorem~\ref{thm:suffCond}.

\begin{comment}
Computing the cutoff for the saturated-model test requires us to know $\proj_{\eta}^\perp Y$, which is not $\sF_k$-measurable.

By contrast, selected-model tests of $M_{k-1}$ against $M_k$ are always $\sF_k$-measurable, because they are based on the law
\begin{equation}\label{eq:selModel}
\L\left(X_{j_k}'Y \mid X_{E_k-1}'Y, E_{k-1}, j_k\right),
\end{equation}
and all of the random variables appearing in~\eqref{eq:selModel} are $\sF_k$-measurable.
\end{comment}

@@

\begin{comment}
First, note that $p_2$ is a selective $p$-value for comparing the null with only variable 1,
\[
M_1:\; Y \sim \cN\left(\binom{\mu_1}{0}, I_2\right),
\]
against the alternative with variables 1 and 2:
\[
M_2:\; Y \sim \cN\left(\mu, I_2\right), \text{ with } 
\mu_2 \neq 0.
\]
To eliminate the nuisance parameter $\mu_1$, the test conditions on $Y_1$. Thus, $p_2$ is uniform given $Y_1$ on $A$. Second, note that on $A$, $p_1$ is a function only of $Y_1$. Thus, $(p_1,p_2)$ are independent uniform variables given that variable 1 is chosen first. By a similar argument, they would also be independent uniforms if $Y_2$ were chosen first; thus, $(p_1, p_2)$ are marginally uniform under the global null.
\end{comment}
